---
title: "Ident Data Cleaning"
author: "Ikedichi Edward Azuh"
format:
 # html:
  #  toc: true
  #  toc-depth: 2
  #  number-sections: true
  #  theme: cosmo
  #  df-print: paged
   pdf:
    #fig-width: 4
    #fig-height: 4.3
    fig-align: center
    fig-show: 'hold'
number-sections: true
toc: false
#editor: source
geometry:
  - top=20mm
  - left=20mm
  - bottom=20mm
  - right=20mm
mainfont: "Times New Roman" 
monofont: "Courier New"      
monofontoptions: 
  - Scale=0.8
execute: 
  cache: true
knitr:
  opts_chunk:
    tidy: false

params:
  hide_answers: false
---

```{r setup, echo=FALSE}
options(dplyr.summarise.inform = FALSE, ggplot2.verbose = FALSE)
```

\frenchspacing

# DATA Pre-processing and cleaning

## Import datasets

### Main data

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(readxl)
library(dplyr)
library(readr)

file_path1 <- "IDENT_2014-2025.xlsx"   # 2014-2025 data
df_2014_2025 <- read_excel(file_path1, sheet = 2, na = c("", "NA", "-"))
head(df_2014_2025)

```


### UTM coordinates data

```{r}

utm_df <- read_excel("IDENT_treeCoords_UTM2.xlsx")
```


# Data cleaning

## Create a simpler unique ID for each Tree using Tree_CN

Sort by Plot, Tree_CN, and Year and generate Tree_ID as a factor using match() and unique() functions.

```{r}
raw_df <- df_2014_2025 %>%
  arrange(Plot, Tree_CN, Year) %>%  # Sort 
  mutate(Tree_ID = as.factor(match(Tree_CN, unique(Tree_CN))))  # Generate Tree_ID 


```

Total number of unique Tree_IDs

```{r}

n_distinct(raw_df$Tree_ID)

```

## Update Height and Diameter

Correct for record error due to height1 and height2 swaps during measurement.

```{r}

####################################
raw_df2 <- raw_df %>%
  # Group by Tree_ID to apply changes within each group
  group_by(Tree_ID) %>%
  mutate(
    # Update Height with Height_2 if Height_2 is greater than Height_1
    Height = ifelse(!is.na(Height_2) & Height_2 > Height_1, Height_2, Height_1),
    
    # Update Diameter with Diameter_2 if Diameter_2 is greater than Diameter_1
    Diameter = ifelse(!is.na(Diameter_2) & Diameter_2 > Diameter_1, Diameter_2,
                      Diameter_1),
    
  ) %>%
  ungroup() %>%
  # Unselect columns like Height_1, Height_2, Diameter_1, etc.
  dplyr::select(-Height_1, -Height_2, -Height_3, -Diameter_1, -Diameter_2, -Diameter_3,
         -Comments, -Tree)


```

## Predict Height measures for Trees where only Diameter was available

Especially from Year=2023 onward.

```{r, warning=FALSE, message=FALSE}

raw_df3a <- raw_df2
raw_df3a$Height <- as.numeric(raw_df3a$Height)
raw_df3a$Diameter <- as.numeric(raw_df3a$Diameter)

# Use only rows where Height is NOT NA (Diameter can be NA)
model_data <- raw_df3a[!is.na(raw_df3a$Height), ]

# Fit a linear model (Height ~ Diameter)
hei_fit <- lm(Height ~ Diameter, data = model_data)
summary(hei_fit)  # Check model output

# Select rows where Height is missing but Diameter is available for prediction
diam_rows <- raw_df3a[is.na(raw_df3a$Height) & !is.na(raw_df3a$Diameter), ]

# Fill NAs with Predicted missing Height using the model
raw_df3a$Height[is.na(raw_df3a$Height) & !is.na(raw_df3a$Diameter)] <- predict(hei_fit, newdata=diam_rows)

# Check if NAs are replaced
sum(is.na(raw_df2$Height)) # NAs not replaced here
sum(is.na(raw_df3a$Height))  # Should be less than before

```

## Now do same for tree where only Height is available

To estimate their corresponding Diameter.

```{r, warning=FALSE, message=FALSE}

#####################################################

raw_df3b <- raw_df3a

# Use only rows where Diameter is NOT NA (Height can be NA)
model_data <- raw_df3b[!is.na(raw_df3b$Diameter), ]

# Fit a linear model 
dia_fit <- lm(Diameter ~ Height, data = model_data)
summary(dia_fit)  # Check model output

# Select rows where Diameter is missing but Height is available
hei_rows <- raw_df3b[is.na(raw_df3b$Diameter) & !is.na(raw_df3b$Height), ]

# Predict missing Diameter using the model
raw_df3b$Diameter[is.na(raw_df3b$Diameter) & !is.na(raw_df3b$Height)] <- predict(dia_fit, newdata=hei_rows)

# Check if NAs are replaced
sum(is.na(raw_df3a$Diameter)) # NAs not replaced here
sum(is.na(raw_df3b$Diameter))  # Should be less than before

```

## More on missing values

Replace "NA" with average of previous and next values if available.

```{r, warning=FALSE, message=FALSE}

raw_df3c <- raw_df3b %>%
  # Group by Tree_ID to apply changes within each group
  group_by(Tree_ID) %>%
  mutate(
    
# Handle missing Height values (replace with average of previous and next values)
    Height = ifelse(is.na(Height), 
                (lag(Height, order_by = Year) + lead(Height, order_by = Year))/2, 
                 Height),
    
  # Handle missing Diameter values (replace with ave. of previous and next values)
    Diameter = ifelse(is.na(Diameter), 
              (lag(Diameter, order_by = Year) + lead(Diameter, order_by = Year))/2, 
                Diameter)
                      
  ) %>%
  # Remove grouping
  ungroup()     

sum(is.na(raw_df3b$Height)) # NAs not replaced here
sum(is.na(raw_df3c$Height))  # Should be less than before
```

## Function to replace values after consecutive NAs with NA

```{r, warning=FALSE, message=FALSE}

# Replace values that followed two or more consecutive 
update_na_values <- function(group) {
  n <- nrow(group)
  heights <- group$Height
  diameters <- group$Diameter
  
  # Find positions of NAs in Height
  na_positions <- which(is.na(heights))
  
  if (length(na_positions) == 0 || length(na_positions) == n) {
    return(group)  # No NA or all NA, return unchanged
  }
  
  # Iterate through the positions of NAs
  for (i in seq_along(na_positions)) {
    pos <- na_positions[i]
    
    if (pos > 1 && pos < n) {  # Ensure it's between values
      prev_value <- heights[pos - 1]
      next_pos <- pos
      
      # Move forward to find the end of the NA sequence
      while (next_pos < n && is.na(heights[next_pos])) {
        next_pos <- next_pos + 1
      }
      
# If next value is non-NA and the previous value is also non-NA, then we have a bounded NA sequence
      if (next_pos <= n && !is.na(heights[next_pos]) && !is.na(prev_value)) {
        # Replace all values after the NA sequence with NA in both Height and Diameter
        heights[next_pos:n] <- NA
        diameters[next_pos:n] <- NA
        break  # Stop after first bounded sequence
      }
    }
  }
  
  group$Height <- heights
  group$Diameter <- diameters
  return(group)
}

# Apply function to each group of Tree_ID
raw_df4 <- raw_df3c %>%
  group_by(Tree_ID) %>%
  group_modify(~ update_na_values(.x)) %>%
  ungroup()

```

## Assign status
Create "Status" column using the following conditions:

If for a tree:
1.  Only first 2 years have values \~ "D"
2.  Only first year has value \~ "C"
3.  Only second, third, or fourth year has a value \~ "Delete"
4.  All entries are NAs \~ "Delete"
5.  Maximum year is NA \~ "D"
6.  Maximum year has values \~ "A"


```{r, warning=FALSE, message=FALSE}
library(dplyr)

  
  # Assign Status
raw_df5 <- raw_df4 %>%
  group_by(Tree_ID) %>%
  arrange(Year, .by_group = TRUE)%>%
  mutate(
    # Count non-NA values
    non_na_count = sum(!is.na(Height)),
    
    # Years with values
    years_with_values = list(Year[!is.na(Height)]),
    
    # First and last year with measurements
    first_year = ifelse(length(years_with_values[[1]]) > 0,
                        min(years_with_values[[1]], na.rm = TRUE), NA),
    last_year = ifelse(length(years_with_values[[1]]) > 0,
                       max(years_with_values[[1]], na.rm = TRUE), NA),
    
    # Only first two years (2014, 2015)
    first_two_only = length(years_with_values[[1]]) == 2 &&
                     all(years_with_values[[1]] %in% c(2014, 2015)),
    
    # Assign Status
    Status = case_when(
      non_na_count == 0 ~ "Delete",
      non_na_count == 1 & first_year %in% c(2015, 2016, 2017, 2018) ~ "Delete",
      non_na_count == 1 & first_year == 2014 ~ "C",
      first_two_only ~ "D",
      last_year == 2025 ~ "A",
      last_year < 2025 ~ "D",
      TRUE ~ NA_character_
    ) %>%
      as.factor()
  ) %>%
  ungroup() %>%
  
  # Remove helper columns
  select(-non_na_count, -years_with_values, -first_year, -last_year, -first_two_only)

table(complete.cases(raw_df5))  # FALSE:34344,  TRUE: 79781 

check = raw_df5 %>%
  filter(Plot == 177, Year >= 2022)
check[, c(11:14)]
```



You can have a look at number of rows to be deleted

```{r}

row_to_delete = raw_df5[raw_df5$Status=="Delete", ]
n_distinct(row_to_delete$Tree_ID)

```

How many Trees were censored?

```{r}
censored_trees = raw_df5[raw_df5$Status=="C", ]
n_distinct(censored_trees$Tree_ID)
```

## More UPDATE: now delete the unwanted rows

```{r}

raw_df6 <- raw_df5 %>%
  filter(Status != "Delete")

```

## Now add the UTM coordinates to the dataset (raw_df7)

```{r}
raw_df7 <- raw_df6 %>%
  left_join(utm_df %>% dplyr::select(Tree_CN, X, Y), by = "Tree_CN")

head(raw_df7)
```

## Create/add Origin, Basal area (BA), and Species richness and Leaftype to dataset

```{r}
############################

raw_df8 <- raw_df7 %>%
  # Create the 'Origin' column and convert to factor
  mutate(
    Origin = case_when(
      Tree_Species %in% c("ACPL", "LADE", "BEPE", "PIAB", "QURO", "PISY") ~ "Native",
      TRUE ~ "Exotic"
    ) %>%
      as.factor(),  # Convert to factor within mutate
    
     Leaftype = case_when(
      Tree_Species %in% c("ACPL", "ACSA", "BEPE", "BEPA", "QURO", "QURU") ~ "Broad",
      TRUE ~ "Needle"
    ) %>% as.factor(),
    
    # Create SR based on the length of Species_Composition
    SR = nchar(as.character(Species_Composition))%>%
      as.factor(),  # Convert to factor,
    
    # Categorize SR as Mono or Multi
    SR_cat = ifelse(SR == 1, "Mono", "Multi") %>%
      as.factor(),  # Convert to factor
    
    # Map 'Year' to numeric values (nYear)
    nYear = case_when(
      Year == 2014 ~ 1,
      Year == 2015 ~ 2,
      Year == 2016 ~ 3,
      Year == 2017 ~ 4,
      Year == 2018 ~ 5,
      Year == 2019 ~ 6,
      Year == 2020 ~ 7,
      Year == 2021 ~ 8,
      Year == 2022 ~ 9,
      Year == 2023 ~ 10,
      Year == 2024 ~ 11,
      Year == 2025 ~ 12
    )
  ) %>%
  # Create Basal Area (BA) in m^2 and impute missing values with median
mutate(BA = pi * (Diameter / 20)^2 / 10000) %>%
mutate(BA = ifelse(is.na(BA), median(BA, na.rm = TRUE), BA))  # Impute missing values with median


```

# Create/add start, stop variables

Using nYear and also recode Status to change at the end of measures for an ID

```{r}

raw_df9 <- raw_df8 %>%
  filter(!is.na(Height)) %>%  # Remove rows where Height is NA
  arrange(Tree_ID, Year) %>%  # Ensure the data is sorted by Tree_ID and Year
  group_by(Tree_ID) %>%
  mutate(
    start = ifelse(is.na(lag(nYear)), nYear - 1, lag(nYear)),  # Set start to the previous year
    stop = nYear,                                           # Set stop to the current year
    Time_1 = n(),                            # Total number of observations for each Tree_ID
    Time_2 = row_number()                              # Sequential count of observations
  ) %>%
  ungroup()

# Recoding the 'Status' column: Change all 'D' to 'A' except for the last occurrence
raw_df9 <- raw_df9 %>%
  group_by(Tree_ID) %>%
  mutate(
    Status = as.character(Status),  # Ensure Status is a character vector
    Status = ifelse(row_number() != n() & Status == "D", "A", Status)
  ) %>%
  ungroup()

# Create binary Status2 column (1 for Dead, 0 for Alive)
raw_df9$Status2 = ifelse(raw_df9$Status == "D", 1, 0)

# Check structure of the dataset
str(raw_df9)

remove(raw_df,raw_df2,raw_df3a,raw_df3b, raw_df4,raw_df3c,raw_df5, raw_df6, raw_df7, 
       raw_df8)
```

# Computing Heygi's competition index (1974)

Here we use diameter as measure of size and competition radius of 1,2 and 3. 

```{r, message=FALSE, warning=FALSE}

library(sf)
library(units)

# Convert the dataset into an sf object using X and Y as coordinates
tree_sf <- st_as_sf(raw_df9, coords = c("X", "Y"), crs = 32633)

# Unique years in the dataset
years <- unique(tree_sf$Year)

# Competition radii to loop over
competition_radii <- c(1, 2, 3)

# Initialize a list to store CI results for each year
ci_results <- list()

# Loop through each year and calculate CI for each radius and each tree
for (year in years) {
  
  # Filter data for the current year
  year_data <- tree_sf %>% filter(Year == year)
  
  # Calculate pairwise distances for the current year
  dist_matrix <- st_distance(year_data)
  dist_matrix <- set_units(dist_matrix, NULL)  # Remove units for easier computation
  
  # Loop through each competition radius
  for (radius in competition_radii) {
    
    # Function to calculate CI for each tree (Heygi's Competition Index)
    calculate_ci <- function(i, diameter, dist_matrix, radius) {
      distances <- dist_matrix[i, ]
      distances <- as.numeric(distances)
      
      # Identify neighbors within the competition radius (excluding the focal tree itself)
      within_radius <- which(distances <= radius & distances > 0)
      
      # Filter out neighbors with missing heights
      valid_neighbors <- within_radius[!is.na(diameter[within_radius])]
      
      if (length(valid_neighbors) == 0) {
        return(0)  # No valid neighbors within radius
      }
      
      # Get heights and distances for valid neighbors
      diameters_within_radius <- diameter[valid_neighbors]
      distances_within_radius <- distances[valid_neighbors]
      
      # Calculate the CI value for the target tree i (HCI)
      ci_value <- sum(diameters_within_radius / (diameter[i] * distances_within_radius))
      return(ci_value)
    }
    
    # Apply CI calculation to each tree for the given radius
    year_data[[paste0("CI_", radius)]] <- sapply(1:nrow(year_data), 
                                                 function(i) calculate_ci(i, year_data$Diameter, 
                                                                          dist_matrix, radius))
  }
  
  # Store results for this year
  ci_results[[as.character(year)]] <- year_data
}

# Combine the results into a single data frame
ci_combined <- do.call(rbind, ci_results)

############################################
# Calculate the median CI for each of the CI columns
############################################
ci_medians <- sapply(competition_radii, function(radius) {
  median(ci_combined[[paste0("CI_", radius)]], na.rm = TRUE)
})

############################################
# Classify each CI column as "Dominant" or "Subordinant"
############################################
for (radius in competition_radii) {
  ci_column <- paste0("CI_", radius)
  cat_column <- paste0(ci_column, "_cat")
  
  ci_combined[[cat_column]] <- ifelse(ci_combined[[ci_column]] <= ci_medians[[radius]],
                                      "Dominant", "Subordinant")
}

df_ci_combined <- as.data.frame(ci_combined)

remove(tree_sf,years,competition_radii,ci_combined,ci_medians,ci_results)

```

## Which value of radius to use?

Fit a logistic regression model

```{r,echo=FALSE, warning=FALSE, message=FALSE}
# Load necessary library
library(pROC)

# Define a function to fit models and compute AIC and AUC
calculate_metrics <- function(data, prefix, variable_prefix) {
  aic_values <- numeric(3)
  auc_values <- numeric(3)

  for (i in 1:3) {
    # Construct the formula without spaces
    formula <- as.formula(paste("Status2 ~", paste(variable_prefix, i, sep = "")))
    model <- glm(formula, data = data, family = binomial)
    
    # Store AIC
    aic_values[i] <- AIC(model)
    
    # Calculate predicted probabilities and AUC
    pred_probs <- predict(model, type = "response")
    roc_curve <- roc(data$Status2, pred_probs)
    auc_values[i] <- auc(roc_curve)
  }
  
  # Combine AIC and AUC into a data frame
  results <- data.frame(Model = paste(prefix, 1:3, sep = "_"),
                         AIC = aic_values,
                         AUC = auc_values)
  return(results)
}

# Calculate metrics for HCI models

hci_metrics <- calculate_metrics(df_ci_combined, "HCI", "CI_")

# Print the combined results
print(hci_metrics)

```

According to the performance metrics used (AIC and AUC), the best radius values for HCI is 3.

# Some exploration

```{r,echo=FALSE, warning=FALSE, message=FALSE}
raw_df10 <- df_ci_combined[, !(names(df_ci_combined) %in% c("CI_1", 
                                                            "CI_1_cat"
                                                            ))]

# we can save this for now as our cleaned data
# write.csv(raw_df10, "cleaned_data.csv")
```

## Plot the tree coordinates

```{r,fig.width=10, fig.height=5, warning=FALSE, message=FALSE}
#| fig.width: 12
#| echo: true
#| eval: true
#| fig.height: 5
#| warning: false
#| fig-cap: UTM coordinates of tree by block and plots.

library(ggplot2)
library(patchwork)
library(sf)


# Convert the dataset into an sf object

tree_sf <- st_as_sf(raw_df10)
                    
plot1 <- ggplot(data = tree_sf) +
  geom_sf(aes(color = Block)) +
  labs(title = "Tree Locations in UTM Coordinates (2013)",
       color = "Block") +
  theme_minimal()

# Filter the data for Block == 1 and Plot == 1
filtered_trees <- tree_sf %>%
  filter(Year == 2025)

# Plot the filtered tree locations as points
plot2 <- ggplot(data = filtered_trees) +
  geom_sf(aes(color = Block)) +
  labs(title = "Tree Locations in UTM Coordinates (2025)") +
  theme_minimal()

# Combine the plots side by side
plot1 + plot2

remove(plot1,plot2)
```

## Competition index vs. height, by tree mortality status

```{r,fig.width=10, fig.height=5,warning=FALSE, message=FALSE}
#| fig.width: 10
#| echo: true
#| eval: true
#| fig.height: 4
#| message: false
#| warning: false
#| fig-cap: Plot of tree height vs competition indices by tree status (a) and competition status (b).
library(patchwork)
library(gridExtra)

plot_data1 <- raw_df10 %>%
  group_by(Tree_ID) %>%
  filter(Year == max(Year)) %>%
  ungroup()


p1 <- ggplot(plot_data1, aes(y = Height, x = CI_3, color = as.factor(Status2))) +
  geom_point() +
  labs(title =  bquote(bold("(a)")),
       y = "Tree Height (cm)", x = "CI",
       color = "Tree Status") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(trans = 'log')  # Apply log scale to x-axis
 # theme(legend.position = "bottom",  # Position legend at the bottom
 #       plot.title = element_text(face = "bold"))

# Customize colors for the legend
custom_colors <- c("0" = "green", "1" = "red")

p1 <- p1 + scale_color_manual(values = custom_colors)


p2 <- ggplot(plot_data1, aes(y = Height, x = CI_3, color = CI_3_cat))  +
  geom_point() +
  labs(title = bquote(bold("(b)")),
       y = "Tree Height (cm)", x = "CI",
       color = "Competition Status:") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), plot.title = element_text(face = "bold")) +
  scale_x_continuous(trans = 'log')  # Apply log scale to x-axis
 # theme(legend.position = "bottom",  # Position legend at the bottom
 #       plot.title = element_text(face = "bold"))


grid.arrange(p1,p2, nrow = 1)
remove(p1,p2)
```

## Mono vs Multi settings of species richness

```{r,fig.width=10, fig.height=8, warning=FALSE, message=FALSE}
# Mono vs Multi settings

ggplot(data=raw_df10, aes(x = factor(Year), y = Height, fill=Origin)) + 
  geom_boxplot() +
  facet_grid(SR_cat ~ CI_3_cat) +
  labs(title = "Tree Height Over Years: Mono vs settings Multi settings of species richness",
       x = "Year",
       y = "Height (cm)",
       color = "Origin") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

```
